---
title: "Independent Project Final Report"
author: "Timara Vereen"
date: "2025-05-14"
output: html_document
---

### Introduction
Vector behavior is a key factor in disease transmission, influencing pathogen spread among humans, wildlife, and domestic animals. Various arthropod vectors, such as mosquitoes, fleas, and tsetse flies, exhibit host-seeking behaviors that shape disease dynamics (Benelli, 2020; Onen et al., 2023; Rogers & Bates, 2007; Pipová et al., 2023). Vector-borne diseases account for over 17% of infectious diseases, leading to more than 700,000 deaths annually (World Health Organization, 2020; Sutherust, 2004; Chala & Hamde, 2021). Climate change, land-use alterations, and human activity are driving the increasing prevalence of these diseases by disrupting ecological systems and increasing host-vector interactions (Ellwanger et al., 2020; Swei et al., 2020). While the behaviors of many vectors have been documented (Azad et al., 1997; Gleave et al., 2016; Bruno et al., 2019; Javed et al., 2021; Stanczyk et al., 2017; Wei Xiang et al.), nymphal tick questing behavior, particularly in Ixodes pacificus, remains less understood, despite its implications for disease transmission.

Ixodes pacificus, the primary vector of Lyme disease (LD) in the western U.S. (CDC, 2020), plays a crucial role in pathogen transmission, with over 400,000 new cases estimated annually in the U.S. (Garira & Chirove, 2019; CDC, 2024). While Ixodes scapularis nymphs in the eastern U.S. exhibit well-documented host-seeking behaviors, I. pacificus questing activity is more variable and less frequently observed, particularly in Southern California (Lane & Kjemtrup, 2016; Lane et al., 2013). Geographic, genetic, and environmental factors likely contribute to these behavioral differences, influencing LD incidence across regions (Arsnoe et al., 2019). Additionally, host interactions, such as those with Western fence lizards (Sceloporus occidentalis), may shape tick activity, as lizard-associated tick abundances fluctuate seasonally (Lane & Loye, 1989). These variations in questing behavior highlight the need for further research to understand the environmental and ecological drivers of tick activity, which could aid in disease prevention and risk assessment (Tomkins et al., 2014).
Our goals are to: 1) Does geographic origin affect I. pacificus questing behavior? I hypothesize that questing behavior will differ based on geographic origin during observation trials. Specific Aim 1: Determine whether geographic origin affects questing behavior due to different environmental conditions in these regions


### Aim 1

Specific Aim 1: Determine whether geographic origin affects questing behavior due to different environmental conditions in these regions

### Hypotheses

Null Hypothesis: Geographic origin has no effect on tick questing behavior, regardless of the environmental conditions associated with different regions.

Alternative Hypothesis: Geographic origin affects tick questing behavior, likely due to differences in environmental conditions between regions.

### Dataset Description

Description: This dataset contains observations of questing behavior in Ixodes pacificus nymphal ticks, focusing on horizontal and vertical movement along with activity levels. The data explores the influence of geographic origin on questing behavior across different locations in California. Each tick was observed over a 5-minute trial, with movement recorded at 30-second intervals. The dataset includes multiple trials per tick and considers factors such as engorgement level and population origin.
https://github.com/vereentn/Tick_Behavior/blob/main/Data/dataset_card.md

### Statistical approach

Behavioral trials were conducted in two different laboratory rooms, each with distinct environmental conditions. The first room, maintained at approximately 17.46°C and 70.18% relative humidity (hereafter referred to as the “17.46°C room”), was located in the San Francisco Animal Care Facility and housed four mus musculus (C3H) mice as part of another experiment. Initial trials were conducted in this space, which offered more consistent temperature and humidity control. For this project, I will be reporting on data from the 17.46°C room.

All statistical analyses were conducted in R (version 4.4.1) using RStudio, with key packages including tidyverse, glmmTMB, performance, and ggplot2 for data cleaning, visualization, and modeling. Questing behavior metrics were analyzed using Generalized Linear Mixed Models (GLMMs) to account for non-normal distributions and repeated measures. Total distance, horizontal distance, and vertical gross distance were modeled using GLMMs with a Tweedie distribution to account for overdispersion and the continuous nature of the distance data. Percent activity, a proportion-based metric with many zero values, was analyzed using a negative binomial GLMM with a zero-inflated component. For the sake of this project, I will be reporting on the total distance traveled behavioral metric. Fixed effects included tick population of origin and engorgement level across all models. Tick ID and trial number were included as random effects to account for repeated observations and individual variation. Model performance was evaluated using Akaike Information Criterion (AIC), with lower AIC values guiding model selection. 
https://github.com/vereentn/Tick_Behavior/tree/main

# Stats and Visuals
```{r}
library(ggplot2)
library(ggsignif)
library(glmmTMB)
library(readr)
Master_s_data_results_4_1_25 <- read_csv("Master's data - results_4.1.25.csv")
#View(Master_s_data_results_4_1_25)
data <- Master_s_data_results_4_1_25
data$Engorgement <- as.factor(data$Engorgement)
data$horizontal_dist_cm <- data$horizontal_dist * 0.635
data$vertical_dist_gross_cm <- data$vertical_dist_gross * 0.635
data$total_dist_cm <- data$total_dist * 0.635
data_u <- subset(data, Room == "U") # U refers to 17.46 room

```


```{r}
#total distance (u)
model_total<- glmmTMB(
  total_dist_cm ~ Population * Engorgement + (1|Tick_ID) + (1|Trial),
  family = tweedie(link = "log"),
  data = data_u
)

summary(model_total)

model_total1<- glmmTMB(
  total_dist_cm ~ Population + Engorgement + (1|Tick_ID) + (1|Trial),
  family = tweedie(link = "log"),
  data = data_u
)

summary(model_total1)

AIC(model_total, model_total1)
#model_total has a lower AIC score, so we used that model (model_total)


```

### Graphs for the 17.46C Room
```{r}
data_u$Pop_Engorgement <- paste(data_u$Population, data_u$Engorgement, sep = "\n")

# Okabe-Ito palette for two groups
okabe_ito <- c("Northern California" = "#E69F00",  # Orange
               "Southern California" = "#56B4E9")  # Sky Blue

ggplot(data_u, aes(x = as.factor(Room), y = total_dist_cm, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, position = position_dodge(0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "black",
               position = position_dodge(0.75)) +
  labs(
    x = "Room",
    y = "Total Distance (cm)",
    fill = "Population",
     caption = "Boxplots represent gross total distance traveled by Ixodes pacificus nymphs.\nData grouped by geographic origin."
  ) +
  theme_minimal() +
  scale_fill_manual(values = okabe_ito) +
  geom_signif(
    comparisons = list(c("U", "U")),  # You may want to check this comparison
    annotations = "p = 0.032*",
    y_position = max(data$total_dist_cm) * 0.8,
    tip_length = 0.08,
    textsize = 4
  )+
  theme(
  plot.title = element_text(face = "bold", hjust = 0.5, size = 18),
  axis.title = element_text(face = "bold", size = 14),
  axis.text = element_text(size = 12),
  axis.text.x = element_text(angle = 45, hjust = 1),
  legend.title = element_text(face = "bold", size = 13),
  legend.text = element_text(size = 11)
)


```


```{r}
# Comparison bracket data
sig_df <- data.frame(
  xmin = c(0.75, 0.75),
  xmax = c(1.2, 2.2),
  y_position = c(27, 36),
  annotation = c("p = 0.048", "p = 0.020")
)

# Plot
ggplot(data_u, aes(x = Population, y = total_dist_cm, fill = as.factor(Engorgement))) +
  geom_boxplot(outlier.shape = NA, width = 0.6, position = position_dodge(width = 0.7)) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 20,
    size = 3,
    color = "black",
    position = position_dodge(width = 0.7)
  ) +
  geom_signif(
    data = sig_df,
    aes(xmin = xmin, xmax = xmax, annotations = annotation, y_position = y_position),
    manual = TRUE,
    inherit.aes = FALSE,
    tip_length = 0.02,
    textsize = 3.5
  ) +
  scale_fill_manual(
    values = c("1" = "#a6cee3", "2" = "#1f78b4", "3" = "#073e85ff"),
    labels = c("1" = "Unfed", "2" = "Partially Engorged", "3" = "Fully Engorged")
  ) +
  labs(
    x = "Population",
    y = "Total Distance (cm)",
    fill = "Engorgement Level",
    caption = "Boxplots represent gross total distance traveled by Ixodes pacificus nymphs.\nData grouped by geographic origin and engorgement level (1 = unfed, 2 = partially engorged, 3 = fully engorged)."
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

```


### Results

Here’s the first main result: when comparing gross total distance traveled — which reflects the overall movement of ticks during the trial — I found a significant difference between populations.
Northern California ticks moved significantly more than Southern California ticks across trials.
This suggests that geographic origin does influence questing behavior — ticks from the north are more active overall, which could translate to increased chances of encountering a host.

When we break this down further by engorgement level, some interesting patterns emerge.
Unfed Northern California ticks moved the most, which makes sense — they’re likely still host-seeking.
 In contrast, Southern California unfed ticks moved much less, suggesting reduced questing behavior.
Surprisingly, fully engorged Southern California ticks moved significantly more — which might reflect post-feeding dispersal or a search for molting sites.
This interaction shows that behavior isn’t just geography or physiology alone — it’s both

### Big Picture 

In thinking about the broader implications of this study, one important question is: do our findings align with known patterns of Lyme disease risk in California?
According to CDPH, Lyme disease incidence is notably higher in Northern California—particularly in coastal and woodland regions where Ixodes pacificus nymphs are more commonly encountered by humans.
Our behavioral data reflects this: we found that Northern California ticks, especially those at early engorgement stages, are significantly more active. This suggests a stronger host-seeking drive, which makes sense biologically—these unfed ticks are actively searching for a blood meal. This elevated activity increases the likelihood of host contact, including with people, and aligns with the elevated Lyme risk in these areas.
In contrast, ticks from Southern California were much less active when unfed. That inactivity may reduce the chance of human-tick encounters, especially during dry seasons—possibly contributing to the lower Lyme disease reports in this region.
However, what's really interesting is that we observed a spike in activity among Southern California ticks after they had fed. This post-feeding movement could be related to dispersal or the search for microhabitats to molt. While these ticks may not pose an immediate transmission risk, their increased movement might still contribute to pathogen spread—especially if they’re dropped off by pets or wildlife, or as climate patterns shift.

So overall, these findings not only support existing public health surveillance data, but also highlight the need to consider behavioral context—when and why ticks move—not just where they are found.

To wrap up, this study shows that tick behavior — plays a critical role in shaping Lyme disease risk.
While Ixodes pacificus is widespread throughout California, my findings suggest that ticks in different regions behave differently, and that these differences align with regional Lyme disease incidence.
Understanding when ticks are active — not just where they are — can help us design more targeted prevention strategies, enhance surveillance efforts, and ultimately reduce Lyme disease risk in California and beyond.

